---
title: "NYPD Shooting Incident Data Analysis"
author: "Ryan Phillips"
date: "2023-06-14"
output: html_document
---

Necessary packages for this analysis: tidyverse core packages.

## Step 0: Import Library

```{r load_tidyverse, message=FALSE}
##install.packages("tidyverse") to install
library(tidyverse)
```

## Step 1: Read in Data

Begin by reading in data from csv file from the City of New York.

```{r import_data, message=FALSE}
NYPD_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

```

## Step 2: Clean and Transform Data

Select only columns relevant to our investigation of this data

```{r tidy_NYPD_data}
NYPD_data <- NYPD_data %>% mutate(OCCUR_DATE= mdy(OCCUR_DATE)) %>% select(-c(JURISDICTION_CODE,INCIDENT_KEY,LOC_OF_OCCUR_DESC,LOC_CLASSFCTN_DESC,STATISTICAL_MURDER_FLAG,Latitude,Longitude,Lon_Lat,X_COORD_CD,Y_COORD_CD,LOCATION_DESC,PRECINCT, PERP_RACE, VIC_RACE)) 


```

Rename Date and time

```{r rename_Date_Time}
NYPD_data <- NYPD_data %>% rename(Date = OCCUR_DATE, Time = OCCUR_TIME)

```

Filter out NA and UNKNOWN variables

```{r drop_na_NYPD_data}
NYPD_data_tidy <- NYPD_data %>% drop_na()
# filter out Unknown vic age group variables
NYPD_data_tidy <- NYPD_data_tidy %>% filter(!grepl('UNKNOWN', VIC_AGE_GROUP))
NYPD_data_tidy <- NYPD_data_tidy %>% filter(!grepl('1022', VIC_AGE_GROUP))

```

Summary of our cleaned data

```{r summary_of_data}
summary(NYPD_data_tidy)
```

## Step 3: Add Analysis and Visualization

Lets sort victim age groups in descending order by total number, to see which age group has the most instances.

```{r order_age_groups}
NYPD_data_tidy %>% group_by(VIC_AGE_GROUP) %>% summarise(Total=n()) %>% arrange(desc(Total))
```

Create a bar chart to visualize instances per age group.

```{r bar_graphs}
vic_age_data <- NYPD_data_tidy %>% group_by(VIC_AGE_GROUP) %>% summarize(incidents = n())

ggplot(vic_age_data, aes(x=VIC_AGE_GROUP, y=incidents, fill=VIC_AGE_GROUP)) +
  geom_bar(stat = "identity") +
  xlab("AGE") + ylab("INCIDENTS") +
  ggtitle("SHOOTING INCIDENTS BY AGE OF VICTIMS")+
  theme_light()
```

I want to see which age group has the highest chance of becoming a victim of a shooting. I begin by creating a population column for each age range.

```{r age_groups_by_population}
NYPD_data_tidy <- NYPD_data_tidy %>% 
  mutate(Population = case_when(
    VIC_AGE_GROUP == "<18" ~ 1798842,
    VIC_AGE_GROUP == "18-24" ~ 773258,
    VIC_AGE_GROUP == "25-44" ~ 2708853,
    VIC_AGE_GROUP == "45-64" ~ 2101599,
    VIC_AGE_GROUP == "65+" ~ 1155075,
    TRUE ~ NA_real_
  ))
```

Next I calculate and plot the shooting incident rate per 100,000 people for each age group.

```{r per_capita_age_group}
vic_age_data_rate <- NYPD_data_tidy %>%
  group_by(VIC_AGE_GROUP) %>%
  summarise(total_shootings = n(),
            population = unique(Population),
            shooting_rate = total_shootings / (population / 100000)) %>%
  arrange(desc(shooting_rate))

ggplot(vic_age_data_rate, aes(x = VIC_AGE_GROUP, y = shooting_rate)) +
  geom_bar(stat = "identity", fill = "turquoise", color = "black") +
  ggtitle("Chance of Getting Shot by Age") +
  xlab("Age") +
  ylab("Shooting Rate per 100,000 People") +
  theme_minimal()
```

Now I would like to see what time of day that shootings are most likely to occur.
```{r time_of_day}
NYPD_data_tidy$HOUR = hour(hms(as.character(NYPD_data_tidy$Time)))

occur_hour = NYPD_data_tidy %>% group_by(HOUR) %>% count()

ggplot(occur_hour, aes(x=HOUR, y=n)) +
  geom_line() +
  labs(title = "What time should people be most cautious of shootings in NYC",
       x = "Hour Occured",
       y= "Number of Incidents") +
  theme_minimal()

```